---
title: "Note on goodness of fit "
format: html
editor: visual
---

Short note on comparing the results of `vcd::goodfit` and `stats::chisq.test`.

```{r}
library(vcd)
```

Data from the lecture. 

```{r}
X <- rep(0:4, c(100,50,15,5,1))
X_tab <- table(X)
X_tab
```

Goodness of fit using the `vcd::goodfit` function, where $\lambda$ is estimated using maximum likelihood (`method = "ML"`) or minimum $\chi^2$ (`method = "MinChisq"`). 

```{r}
res_ml <- goodfit(X, "poisson", method = "ML")
res_x2 <- goodfit(X, "poisson", method = "MinChisq")
```

Note that $\hat{\lambda}$ is estimated differently so we get: `r res_ml$par` for the `ML` and `r res_x2$par` for the `MinChisq` method. 

The first object returns $G^2$ statistics and tests instead of $\chi^2$ statistics and tests.

```{r}
summary(res_ml)
```

The second object returns $\chi^2$ statistic and test, but the parameters are not estimated by maximum likelihood! 

```{r}
summary(res_x2)
```

If we want to get standard $\chi^2$ statistics using the ML approach, we need to use the `chisq.test` function with the parameter `p` referring to the probabilities from the reference distribution. 

```{r}
chisq.test(table(X), p = res_ml$fitted/sum(res_ml$fitted))
```

Note that the `p-value` is incorrect as `df=4` not `df=3` (correct).
