---
title: "CDA: case study"
format: 
  html:
    self-contained: true
    table-of-contents: true
    number-sections: true
    df-print: kable
editor: visual
execute: 
  eval: true
  warning: false
  message: false
  echo: false
---

```{r}
library(readxl)
```

# Introduction

The aim of the case study is to analyse days of absenteeism (`daysabs`) of 316 students from two Los Angeles high schools (`school`) using their characteristics such as gender (`gender`), ethnicity (`ethnic`), and standardised math (`mathnce`) and language (`langnce`) scores. Example rows from the dataset are presented below


```{r}
df <- read_excel("school-absence-for-lecture.xlsx")
head(df)
```


# Reserch questions and goals

The research questions are as follows:

1. what is the appropriate distribution/model of absentee days? 
2. do males and females differ in terms of absentee days?
3. is there a relationship between standardised test scores and student absenteeism? 
4. if we include all the variables in the model, can we identify the most important one?

In order to answer the research questions the following goals were specified:

1. read the Excel file
2. visualise `daysabs` to assess the distribution and whether zero inflation occurs,
3. plot `daysabs` by `gender`, `ethnicity`, `school`, `mathnce` and `langnce` to assess the relationship,
4. fit count regression models (standard and zero-inflated),
5. evaluate the fit and select an appropriate model
6. explain the model and parameters
7. find important variables using the `Car::Anova` function.






